---
author: Jerry Cummings
title: "Impact of U.S. Extreme Weather Events - 1994 to 2011"
output: html_document
---

Impact of U.S. Extreme Weather Events - 1994 to 2011
========================================================

Raw data from the National Climatic Data Center (NCDC)
concerning weather events between 1950 and Nov. 2011 is
analyzed for two topics of interest:

1. Across the United States, which types of events
   are most harmful with respect to population health?

2. Across the United States, which types of events have
   the greatest economic consequences?

The data volume appears to be rather spotty before 1994, so
our analysis focuses on data since 1-JAN-1994, or more than
seventeen years worth. We find that:

1. Tornadoes have outpaced every other weather event in
   terms of fatalities and injuries during that time period,
   with Heat the next most harmful.
   
2. Floods, Hurricanes and Tornadoes have been the most
   impactful events in terms of damage as measured in U.S. dollars.

Jerry Cummings  
24-MAY-2014
 





<HR>
  
Data Processing
-----------------------------------------------------------------
```{r libraries}
library(stringr)
library(data.table)
library(scales)
library(ggplot2)
library(gridExtra)
```
  
  
```{r get-raw-data, cache=TRUE}
# read the data - read.csv can handle bz2 files.
raw.data <- read.csv(file = 'repdata-data-StormData.csv.bz2',
                     strip.white = TRUE)
```
  
  
```{r examine-raw-data, echo=FALSE}
# Q: which columns have NA values?
#
#sapply(raw.data, function(x) ifelse(sum(is.na(x)) > 0, x, 0))
# A: 'COUNTYENDN', 'F', 'LATITUDE', 'LATITUDE_E'

# what about EVTYPE? seems like some are the same but for caps
#length(unique(raw.data$EVTYPE))          # 985
#length(unique(toupper(raw.data$EVTYPE))) # 898
#
#What are our state values?
#unique(data$STATE)

# Some of those are states, some appear to be territories, and
# some seem to be either fabricated on-the-fly or mistakes. I'm 
# hesitant to change any for this analysis; we'll leave them
# alone. Weird examples:
#
#   PZ? maybe that's for Panama Canal Zone which is now obsolete.
#   GM? maybe that's for Guam, which is GU?
#   LH? who knows?
```
  
  
Make a copy of raw data with a name that reflects that
it is NOT raw data.
```{r copy-raw-data, echo=TRUE}
data <- raw.data
rm(raw.data)
```
  
  
Convert date columns to a Date datatype
```{r date-type-conversions}
data$BGN_DATE <- as.Date(data$BGN_DATE, format='%m/%d/%Y')
data$END_DATE <- as.Date(data$END_DATE, format='%m/%d/%Y')
data$Year <-as.numeric(format(data$BGN_DATE,'%Y'))
```
  
  
What does our weather event distribution look like? Start by
getting a count for each BGN_DATE.
```{r get-date-counts}
DT <- data.table(data)
results <- DT[, length(BGN_DATE), by = Year]
setnames(results, c('Year', 'Count'))
```
  
  
Now plot it. After the first time we ran this, we noted
that the data took a turn for the better in 1994, so let's
annotate that.
```{r plot-date-counts}
p <- ggplot(results, aes(x=Year, y=Count)) +
  geom_point(color='blue') +
  geom_vline(x=1994, color='red') +
  annotate('text', x=1991, y=55000, label='1994', color='red') +
  ggtitle('Number of Weather Events Recorded Per Year') +
  scale_y_continuous(labels = comma) 
print(p)
```
  
  
Move to event types. Event types in the data are all over the place. Let's start with
the valid event types from the document. We'll just cut/paste from
the PDF and then surround it with code.
```{r setup-valid-event-types}
valid.event.types <- c(
  'Astronomical Low Tide',
  'Avalanche',
  'Blizzard',
  'Coastal Flood',
  'Cold/Wind Chill',
  'Debris Flow',
  'Dense Fog',
  'Dense Smoke',
  'Drought',
  'Dust Devil',
  'Dust Storm',
  'Excessive Heat',
  'Extreme Cold/Wind Chill',
  'Flash Flood',
  'Flood',
  'Frost/Freeze',
  'Funnel Cloud',
  'Freezing Fog',
  'Hail',
  'Heat',
  'Heavy Rain',
  'Heavy Snow',
  'High Surf',
  'High Wind',
  'Hurricane (Typhoon)',
  'Ice Storm',
  'Lake-Effect Snow',
  'Lakeshore Flood',
  'Lightning',
  'Marine Hail',
  'Marine High Wind',
  'Marine Strong Wind',
  'Marine Thunderstorm Wind',
  'Rip Current',
  'Seiche',
  'Sleet',
  'Storm Surge/Tide',
  'Strong Wind',
  'Thunderstorm Wind',
  'Tornado',
  'Tropical Depression',
  'Tropical Storm',
  'Tsunami',
  'Volcanic Ash',
  'Waterspout',
  'Wildfire',
  'Winter Storm',
  'Winter Weather'
)
valid.event.types <- toupper(valid.event.types) # convenience
```
  
  
As we said, the raw data is chock full of messiness. Those field guys
need better data entry software so that a bunch of unnecessary hooey
isn't flooding our data. (Ha, ha. Get it? 'flooding'???)
  
Let's ease our data cleanup effort a bit. For now, we'll only look
at events that have FATALITIES, INJURIES, PROPDMG, CROPDMG as those
are the areas that we'll be focusing on in this analysis.
```{r get-subset-of-data}
data <- subset(data, !is.na(FATALITIES) |
                 !is.na(INJURIES) |
                 !is.na(PROPDMG) |
                 !is.na(CROPDMG)
               )
```
  
  
Make a new copy of the event types that we can muck
around with. Go ahead and flip it to upper case and
trim off leading/trailing whitespace.
```{r event-type-cleanup-1}
data$EVTYPE2 <- toupper(str_trim(data$EVTYPE))
```
  
  
Let's do some fixup. Note: order matters in this chunk of code!
```{r event-type-cleanup-2}
data$EVTYPE2[grepl('.*AVAL.*', data$EVTYPE2)] <- 'AVALANCHE'

data$EVTYPE2[grepl('.*T.*WIND.*', data$EVTYPE2)] <- 'THUNDERSTORM WIND'

data$EVTYPE2[grepl('HIGH WIND.*', data$EVTYPE2)] <- 'HIGH WIND'
data$EVTYPE2[grepl('WIND.*', data$EVTYPE2)] <- 'HIGH WIND'
data$EVTYPE2[grepl('.*MICROBURST.*', data$EVTYPE2)] <- 'HIGH WIND'
data$EVTYPE2[grepl('.*MICOBURST.*', data$EVTYPE2)] <- 'HIGH WIND'
data$EVTYPE2[grepl('DOWNBURST.*', data$EVTYPE2)] <- 'HIGH WIND'
data$EVTYPE2[grepl('.*WND.*', data$EVTYPE2)] <- 'HIGH WIND'

data$EVTYPE2[grepl('.*TORN.*', data$EVTYPE2)] <- 'TORNADO'
data$EVTYPE2[grepl('.*NADO.*', data$EVTYPE2)] <- 'TORNADO'

data$EVTYPE2[grepl('FUN.*', data$EVTYPE2)] <- 'FUNNEL CLOUD'
data$EVTYPE2[grepl('.*WALL.*', data$EVTYPE2)] <- 'FUNNEL CLOUD'
data$EVTYPE2[grepl('.*SPOUT.*', data$EVTYPE2)] <- 'WATERSPOUT'

data$EVTYPE2[grepl('HAIL.*', data$EVTYPE2)] <- 'HAIL'

data$EVTYPE2[grepl('.*LIGHTN.*', data$EVTYPE2)] <- 'LIGHTNING'
data$EVTYPE2[grepl('.*LIGHTI.*', data$EVTYPE2)] <- 'LIGHTNING'
data$EVTYPE2[grepl('.*LIGNTN.*', data$EVTYPE2)] <- 'LIGHTNING'

data$EVTYPE2[grepl('.*FLASH.*', data$EVTYPE2)] <- 'FLASH FLOOD'
data$EVTYPE2[grepl('.*DAM.*', data$EVTYPE2)] <- 'FLASH FLOOD'

data$EVTYPE2[grepl('.*FLOOD.*', data$EVTYPE2) &
               !grepl('.*FLASH.*', data$EVTYPE2)] <- 'FLOOD'
data$EVTYPE2[grepl('.*URBAN.*SMALL.*', data$EVTYPE2)] <- 'FLOOD'
data$EVTYPE2[grepl('.*URBAN.*SML.*', data$EVTYPE2)] <- 'FLOOD'
data$EVTYPE2[grepl('.*SMALL.*STR.*', data$EVTYPE2)] <- 'FLOOD'
data$EVTYPE2[grepl('.*RAPID.*RIS.*.*WAT.*', data$EVTYPE2)] <- 'FLOOD'
data$EVTYPE2[grepl('.*SML.*STR.*', data$EVTYPE2)] <- 'FLOOD'

data$EVTYPE2[grepl('.*HURR.*', data$EVTYPE2)] <- 'HURRICANE (TYPHOON)'
data$EVTYPE2[grepl('.*TYPH.*', data$EVTYPE2)] <- 'HURRICANE (TYPHOON)'
data$EVTYPE2[grepl('.*TROP.*STORM.*', data$EVTYPE2)] <- 'TROPICAL STORM'

data$EVTYPE2[grepl('.*HEAVY.*RAIN.*', data$EVTYPE2)] <- 'HEAVY RAIN'
data$EVTYPE2[grepl('.*HEAVY.*PREC.*', data$EVTYPE2)] <- 'HEAVY RAIN'
data$EVTYPE2[grepl('.*REC.*RAIN.*', data$EVTYPE2)] <- 'HEAVY RAIN'
data$EVTYPE2[grepl('RAIN', data$EVTYPE2)] <- 'HEAVY RAIN'

data$EVTYPE2[grepl('.*THUNDER.*', data$EVTYPE2)] <- 'THUNDERSTORM WIND'

data$EVTYPE2[grepl('.*EXTREME COLD.*', data$EVTYPE2)] <- 'EXTREME COLD/WIND CHILL'
data$EVTYPE2[grepl('.*RECORD COLD.*', data$EVTYPE2)] <- 'EXTREME COLD/WIND CHILL'
data$EVTYPE2[grepl('.*RECORD LOW.*', data$EVTYPE2)] <- 'EXTREME COLD/WIND CHILL'
data$EVTYPE2[grepl('.*LOW TEMP.*', data$EVTYPE2)] <- 'EXTREME COLD/WIND CHILL'

data$EVTYPE2[grepl('.*COLD*', data$EVTYPE2)] <- 'COLD/WIND CHILL'

data$EVTYPE2[grepl('.*BLIZZ.*', data$EVTYPE2)] <- 'BLIZZARD'
data$EVTYPE2[grepl('HEAVY SN*', data$EVTYPE2) &
               grepl('.*HIGH*', data$EVTYPE2)] <- 'BLIZZARD'

data$EVTYPE2[grepl('.*FREEZING R.*', data$EVTYPE2)] <- 'WINTER WEATHER'
data$EVTYPE2[grepl('.*SLEET.*', data$EVTYPE2)] <- 'WINTER WEATHER'
data$EVTYPE2[grepl('.*SNOW.*', data$EVTYPE2)] <- 'WINTER WEATHER'
data$EVTYPE2[grepl('.*WINT.*STOR.*', data$EVTYPE2)] <- 'WINTER WEATHER'
data$EVTYPE2[grepl('.*WINT.*MIX.*', data$EVTYPE2)] <- 'WINTER WEATHER'
data$EVTYPE2[grepl('.*FREEZ.*DRIZ.*', data$EVTYPE2)] <- 'WINTER WEATHER'

data$EVTYPE2[grepl('.*ICE.*', data$EVTYPE2)] <- 'ICE STORM'
data$EVTYPE2[grepl('.*GLAZ.*', data$EVTYPE2)] <- 'ICE STORM'

data$EVTYPE2[grepl('.*FREEZE.*', data$EVTYPE2)] <- 'FROST/FREEZE'
data$EVTYPE2[grepl('.*FROST.*', data$EVTYPE2)] <- 'FROST/FREEZE'

data$EVTYPE2[grepl('.*RECORD HEAT.*', data$EVTYPE2)] <- 'EXCESSIVE HEAT'
data$EVTYPE2[grepl('.*RECORD WARM.*', data$EVTYPE2)] <- 'EXCESSIVE HEAT'
data$EVTYPE2[grepl('.*RECORD HIGH.*', data$EVTYPE2)] <- 'EXCESSIVE HEAT'
data$EVTYPE2[grepl('.*HIGH TEMP.*', data$EVTYPE2)] <- 'EXCESSIVE HEAT'

data$EVTYPE2[grepl('.*HEAT.*', data$EVTYPE2) &
               !grepl('.*RECORD H.*', data$EVTYPE2)] <- 'HEAT'
data$EVTYPE2[grepl('.*WARM.*', data$EVTYPE2)] <- 'HEAT'
data$EVTYPE2[grepl('.*HOT.*WEATH.*', data$EVTYPE2)] <- 'HEAT'

data$EVTYPE2[grepl('.*HIGH.*TIDE.*', data$EVTYPE2)] <- 'STORM SURGE/TIDE'
data$EVTYPE2[grepl('.*STORM.*SURG.*', data$EVTYPE2)] <- 'STORM SURGE/TIDE'
data$EVTYPE2[grepl('.*COAST.*SURG.*', data$EVTYPE2)] <- 'STORM SURGE/TIDE'

data$EVTYPE2[grepl('.*HIGH.*SEA.*', data$EVTYPE2)] <- 'HIGH SURF'
data$EVTYPE2[grepl('.*HIGH.*WAV.*', data$EVTYPE2)] <- 'HIGH SURF'
data$EVTYPE2[grepl('.*HEAV.*SEA.*', data$EVTYPE2)] <- 'HIGH SURF'
data$EVTYPE2[grepl('.*HEAV.*SURF.*', data$EVTYPE2)] <- 'HIGH SURF'
data$EVTYPE2[grepl('.*HAZ.*SURF.*', data$EVTYPE2)] <- 'HIGH SURF'

data$EVTYPE2[grepl('.*DUST.*', data$EVTYPE2)] <- 'DUST STORM'
data$EVTYPE2[grepl('.*BLOW.*DUST.*', data$EVTYPE2)] <- 'DUST STORM'

data$EVTYPE2[grepl('.*FIRE.*', data$EVTYPE2)] <- 'WILDFIRE'

data$EVTYPE2[grepl('.*VOLC.*', data$EVTYPE2)] <- 'VOLCANIC ASH'

# During development, report on what's left that didn't
# get mapped to the known types
#x <- unique(data$EVTYPE2[!(data$EVTYPE2 %in% valid.event.types)])
#length(x)
#x

# Just collect those into a new event type that we'll call 'OTHER'
data$EVTYPE2[!(data$EVTYPE2 %in% valid.event.types)] <- 'OTHER'
```
  
  
--- End of Data Processing ---
  
  
  
  
  
  
<HR>
  
Results
--------------------------------------
  
## Fatalities and Injuries due to Extreme Weather Events

Across the United States, which types of events (as indicated
in the EVTYPE/EVTYPE2 variable) are most harmful with respect to
population health?
  
As noted above, for data coverage reasons, we'll only
look at data from 1994 onward.
```{r population-harm-date-subset}
DT <- data.table(subset(data, data$BGN_DATE >= '1994-01-01'))
```
  
  
Aggregate harm to people by event type. There are too many
event types and many aren't impactful, so we'll focus on the top 10.
```{r population-harm-aggregation}
results <- DT[, sum(FATALITIES + INJURIES), by = EVTYPE2]
results <- results[order(V1, na.last=TRUE, decreasing=TRUE),]
top10.event.types <- head(results$EVTYPE2, 10)
DT <- DT[DT$EVTYPE2 %in% top10.event.types,]
results <- DT[, sum(FATALITIES + INJURIES), by = c('EVTYPE2','Year')]
```
  
  
Plot the top event types that harmed the population. Lazy and stupid so brute
forcing it.
```{r population-harm-plot, fig.width=10, fig.height=18}
# starting plot
p1994 <- ggplot(subset(results, Year==1994), aes(x=reorder(EVTYPE2,V1), y=V1)) +
  geom_bar(stat='identity', fill='blue') +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  xlab('Weather Event') +
  ylab('Number Instances') +
  facet_wrap(~Year)

# rest of the plots
p1995 <- p1994 %+% subset(results, Year==1995)
p1996 <- p1995 %+% subset(results, Year==1996)
p1997 <- p1996 %+% subset(results, Year==1997)
p1998 <- p1997 %+% subset(results, Year==1998)
p1999 <- p1998 %+% subset(results, Year==1999)
p2000 <- p1999 %+% subset(results, Year==2000)
p2001 <- p2000 %+% subset(results, Year==2001)
p2002 <- p2001 %+% subset(results, Year==2002)
p2003 <- p2002 %+% subset(results, Year==2003)
p2004 <- p2003 %+% subset(results, Year==2004)
p2005 <- p2004 %+% subset(results, Year==2005)
p2006 <- p2005 %+% subset(results, Year==2006)
p2007 <- p2006 %+% subset(results, Year==2007)
p2008 <- p2007 %+% subset(results, Year==2008)
p2009 <- p2008 %+% subset(results, Year==2009)
p2010 <- p2009 %+% subset(results, Year==2010)
p2011 <- p2010 %+% subset(results, Year==2011)

# output
title <- 'Top 10 United States Weather Event Types for Population Harm\nFatalities and Injuries\n1994 - Nov. 2011'
grid.arrange(p1994, p1995, p1996, p1997, p1998, p1999, p2000,
             p2001, p2002, p2003, p2004, p2005, p2006, p2007,
             p2008, p2009, p2010, p2011,
             nrow = 6, main=title)

# release memory
rm(p1994, p1995, p1996, p1997, p1998, p1999,
   p2000, p2001, p2002, p2003, p2004, p2005,
   p2006, p2007, p2008, p2009, p2010, p2011)
```
  
  
## Economic Impact of Extreme Weather Events
Across the United States, which types of events have the
greatest economic consequences?
  
For economic impact, we care about property damage and
crop damage, which can be found in the following fields:  
   PROPDMG PROPDMGEXP CROPDMG CROPDMGEXP 
  
Regarding the EXP fields, they should contain a value such as K, M or B. “K” for thousands, “M” for millions, and “B” for billions. We need to compute a calculate dollar amount using the exponent fields. To start, we need a vectorized function to translate the "EXP" character
into a multiplier value.
  
```{r multipler}
# > unique(data$CROPDMGEXP)
#  [1]   M K m B ? 0 k 2
# Levels:  0 2 ? B K M k m
#
# > unique(data$PROPDMGEXP)
#  [1] K M   B m + 0 5 6 ? 4 2 3 h 7 H - 1 8
# Levels:  + - 0 1 2 3 4 5 6 7 8 ? B H K M h m

get.multiplier <- function(ch) {
  return(
    switch(EXPR = ch,
           K = 1000,
           M = 1000000,
           B = 1000000000,
           1
           )
  )
}
v.get.multiplier <- Vectorize(get.multiplier)
```

Some cleanup of the EXP fields for sanity...
```{r exp-cleanup}
data$PROPDMGEXP <- toupper(str_trim(data$PROPDMGEXP))
data$CROPDMGEXP <- toupper(str_trim(data$CROPDMGEXP))
```

With clean(ish) EXP values, we can calculate the actual amount
of damage.
```{r calc-damage}
data$PROPDMG_CALC <- data$PROPDMG * v.get.multiplier(data$PROPDMGEXP)
data$CROPDMG_CALC <- data$CROPDMG * v.get.multiplier(data$CROPDMGEXP)
```

1994 is when data reporting began to be at a reasonable volume, so start with data
from then onward.
```{r subset-damage-data}
rm(DT)
DT <- data.table(subset(data, data$BGN_DATE >= '1994-01-01'))
```
  
Aggregate damage by event type.
```{r aggregate-damage}
results <- DT[, sum(PROPDMG_CALC + CROPDMG_CALC), by = EVTYPE2]
results <- results[order(V1, na.last=TRUE, decreasing=TRUE),]
top10.event.types <- head(results$EVTYPE2, 10)
DT <- DT[DT$EVTYPE2 %in% top10.event.types,]
results <- DT[, sum(PROPDMG_CALC + CROPDMG_CALC), by = c('EVTYPE2','Year')]
```

Now plot the top ten.
```{r plot-damage, fig.width=10, fig.height=18}
# starting plot
p1994 <- ggplot(subset(results, Year==1994), aes(x=reorder(EVTYPE2,V1), y=V1/1000000000)) +
  geom_bar(stat='identity', fill='blue') +
  coord_flip() +
  scale_y_continuous(labels = comma) +
  xlab('Weather Event') +
  ylab('Damage - Billions') +
  facet_wrap(~Year)

# rest of the plots
p1995 <- p1994 %+% subset(results, Year==1995)
p1996 <- p1995 %+% subset(results, Year==1996)
p1997 <- p1996 %+% subset(results, Year==1997)
p1998 <- p1997 %+% subset(results, Year==1998)
p1999 <- p1998 %+% subset(results, Year==1999)
p2000 <- p1999 %+% subset(results, Year==2000)
p2001 <- p2000 %+% subset(results, Year==2001)
p2002 <- p2001 %+% subset(results, Year==2002)
p2003 <- p2002 %+% subset(results, Year==2003)
p2004 <- p2003 %+% subset(results, Year==2004)
p2005 <- p2004 %+% subset(results, Year==2005)
p2006 <- p2005 %+% subset(results, Year==2006)
p2007 <- p2006 %+% subset(results, Year==2007)
p2008 <- p2007 %+% subset(results, Year==2008)
p2009 <- p2008 %+% subset(results, Year==2009)
p2010 <- p2009 %+% subset(results, Year==2010)
p2011 <- p2010 %+% subset(results, Year==2011)

# output
title <- 'Top 10 United States Weather Event Types for Economic Impact\nProperty and Crop Damage (Billions USD)\n1994 - Nov. 2011'
grid.arrange(p1994, p1995, p1996, p1997, p1998, p1999, p2000,
             p2001, p2002, p2003, p2004, p2005, p2006, p2007,
             p2008, p2009, p2010, p2011,
             nrow = 6, main=title)

# release memory
rm(p1994, p1995, p1996, p1997, p1998, p1999,
   p2000, p2001, p2002, p2003, p2004, p2005,
   p2006, p2007, p2008, p2009, p2010, p2011)
```


